### 2023-11-20 Manuscript/Draft3.tex
new version of paper

\hypertarget{how-might-climate-change-affect-indices-of-abundance-a-simulation-study-starting-point.}{%
\section{How might climate change affect indices of abundance? A simulation study starting point.}\label{how-might-climate-change-affect-indices-of-abundance-a-simulation-study-starting-point.}}

@@

\textsuperscript{1}Ocean Associates Inc.~in support of \textsuperscript{2}National Marine Fisheries Service, Northeast Fisheries Science Center, 166 Water Street, Woods Hole, Massachusetts 02543-1026

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. At the same time a number of recent studies have found a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area or outside of the assumed stock boundary is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of biomass estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.

@@

\caption{Strata inhabited by each species in our population models. All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange (Walsh et al. 2015). Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row. Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow.}\label{fig:strata-plot}
\end{figure}

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al. 2020). Parameters in the population model account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results. A full list of parameters used in our population model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.

@@

\caption{Visual representation of niche model for Atlantic Cod. In the left most plot, tows with positive Cod catch are shown in red while tows without Cod are in blue. Positive tows are weighted by the biomass of the given catch. The middle two plots show our covariate predictors used in the \textit{lrren} niche model- depth and median sediment size. The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1.}\label{fig:hab-plot1}
\end{figure}

@@

Each species is assumed to have normally distributed temperature preferences \(N(\mu,\sigma)\), with mean \(\mu\) and standard deviation \(\sigma\). Temperature values are in degrees Celcius. Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our population models. \textbf{dont forget to include animated gif in final submission}

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50 years while Haddock has increased in abundance in recent years ((NEFSC) 2022). For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To consider a wide range of possibilities, we include increasing, decreasing, and constant population trends for Yellowtail. The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial data sets.

@@

After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year. We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 and 14 in the spring and 37 and 38 in the fall). The number of simulated tows made reflects true target values for each stratum and sample cells where ``tows'' are made were randomly selected.

An underlying assumption in all index methods is that individual random samples combine to accurately represent true biomass by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.35. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that SRS will fail to sample the entire geographic extent of the population. We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how the distribution of each species evolved in our increasing temperature scenarios and removing strata where the proportion of a given population was changing over time. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species in this scenario. We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.

@@

Knowing the true population values in our simulations allows us to compute the relative absolute error in a given season \(E_s\) from each estimation method and scenario as given by

@@

Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial found at \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}. After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T 2022). Both settings use a Tweedie model and the difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). The specifications for Settings A are more typical for stock assessment application (Thorson, James T 2022), and we therefore focus our discussion on these VAST results. Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.

Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, and covariates), resulting in a large number of scenario combinations to consider. The columns in Table \ref{tab:scenarios} show the choices that define each scenario.

@@

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our 20-year biomass estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). The difference in variance is further illustrated by the median error values, with model-based estimates producing a median value of 0.235 while stratified mean values had a median error of 0.34. When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall absolute errors and standard deviations. Models with covariates had an average absolute error of 0.22 (median 0.19), compared to models that did not include covariate information having an average error of 0.46 (median 0.31), and stratified mean estimates produced an average error of 0.38 (median 0.34). These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true biomass value, while the stratified mean estimates are, on average, below the true biomass values. This can be further examined through yearly estimate:true biomass ratio values, where we divide the yearly biomass estimates by the true biomass value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).

Analysis of individual yearly estimate:true biomass ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed similar patterns.

When the entire domain is sampled, the stratified mean produced an average yearly estimate:true biomass ratio of 0.87 with a standard deviation of 0.12. As the number of years estimated approaches infinity this value should converge to 1. Adding covariates in these scenarios brings the VAST estimate ratio closer to 1. Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in turn decreases the corresponding estimate ratio. For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10). With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19. As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.

@@

Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \textbackslash and B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in \(67/96\approx 70\%\) of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved biomass estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.

@@

Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true biomass trend. While VAST estimates tended to provide lower absolute error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 biomass estimates with larger error values than the largest stratified mean error. These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses. Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken in determining which diagnostic tools are used. For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model (Christopher Cacciapaglia 2023). The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data to account for factors such as environmental change. However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.

Seasonal differences in spatial biomass patterns in our population models led to seasonal differences in biomass estimates, with most estimates producing a higher error in the fall. This may have to do with the fact that our species spawn during weeks 9-14, which coincides with the spring sampling weeks 13 and 14. Spawning in our population models concentrates much of each species in the spawning grounds towards the center of the domain where more samples occur. The populations are more spread out during the fall survey in weeks 37 and 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled. We explored this through biomass estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.

@@

Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (e.g., 0.11 vs 0.13). The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. This is in contrast to our Yellowtail results, where adding covariate information to VAST models always decreased the absolute error in the resulting biomass estimates. Although strata were excluded for both species, the decline in performance for just Cod and Haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists. We notice in Figure \ref{fig:strata-plot} that the strata excluded from sampling for Yellowtail still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate. On the other hand, the specific strata that were excluded from sampling for Haddock were located in the northern region of the domain, which means the covariate response for Haddock was not informed by any data from the coldest temperature range. This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also confounds the ability to project estimates into an unsampled region. Figure \ref{fig:CovResponse} depicts the impact on biomass estimates of including a covariate response for Haddock when both the average temperature and biomass were increasing over time. When projecting into the unsampled strata as seen Figure \ref{fig:CovResponse}, projecting biomass into the unsampled strata provides the wrong magnitude of biomass and also a different trend.

@@

The authors would like to thank Joe Langan for helping to develop the proposal that funded this work. Funding was provided by the Northeast Fisheries and Climate program at the Northeast Fisheries Science Center. Thank you to Jim Thorson for consulting with us about VAST settings and implementation. We thank Chris Cacciapaglia and Charles Adams for the time they spent with us sharing their experiences using VAST. We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ChrisC}{}}%
Christopher Cacciapaglia, C.F.A., Elizabeth N. Brooks. 2023. Developing workflow and diagnostics for model selection of a vector autoregressive spatiotemporal (VAST) model. Fisheries Research \textbf{In Press}. Elsevier.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ManagTrack}{}}%
(NEFSC), N.F.S.C. 2022. Management track assessments fall 2022. NOAA technical memorandum NMFS-NE ; 305.

@@

\leavevmode\vadjust pre{\hypertarget{ref-walsh2015long}{}}%
Walsh, H.J., Richardson, D.E., Marancik, K.E., and Hare, J.A. 2015. Long-term changes in the distributions of larval and adult fish in the northeast US shelf ecosystem. PLoS One \textbf{10}(9): e0137382. Public Library of Science San Francisco, CA USA.

@@

Table 1. Parameters used in all population models.

>>>>

\hypertarget{estimating-population-trends-under-the-pressures-of-climate-change}{%
\section{Estimating Population Trends Under the Pressures of Climate Change}\label{estimating-population-trends-under-the-pressures-of-climate-change}}

@@

\textsuperscript{1}Ben's Institution, USA\\
\textsuperscript{2}National Marine Fisheries Service, Northeast Fisheries Science Center, Woods Hole, MA, USA

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. At the same time a number of recent studies have found a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of biomass estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.

@@

\caption{Strata inhabited by each species in our population models. All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange [@walsh2015long] **fix cite**. Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row. Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow.}\label{fig:strata-plot}
\end{figure}

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al. 2020). Parameters in the population model account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results. A full list of parameters used in our model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.

@@

\caption{Visual representation of niche model for Atlantic Cod. In the left most plot, tows with positive Cod catch are shown in red while tows without Cod are in blue. Positive tows are weighted by the biomass of the given catch. The middle two plots show our covariate predictors used in the *lrren* niche model- depth and median sediment size. The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1.}\label{fig:hab-plot1}
\end{figure}

@@

Each species is assumed to have normally distributed temperature preferences \(N(\mu,\sigma)\), with mean \(\mu\) and standard deviation \(\sigma\). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include animated gif in final submission}

{\centering \includegraphics[width=0.95\linewidth]{Images/TempScenarios}

\caption{Mean trends of temperature data used in our model.}\label{fig:temp-scenarios}
\end{figure}

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}. For this reason we compare survey index estimates using SRS from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To consider a wide range of possibilities, we include increasing, decreasing, and constant population trends for Yellowtail. The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial data sets.

@@

After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting SRS in the set of strata associated with each stock twice each year. We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 \& 14 in the spring and 37 \& 38 in the fall). The number of simulated tows made reflects true target values for each stratum and sample cells where ``tows'' are made were randomly selected.

An underlying assumption in all index methods is that individual random samples combine to accurately represent true biomass by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.35. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that SRS will fail to sample the entire geographic extent of the population. We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how the distribution of each species evolved in our increasing temperature scenarios and removing strata where the proportion of a given population was changing over time. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species. We then use the biomass collected from our samples in contemporary biomass index methods to estimate yearly seasonal population trends.

@@

Knowing the true population values in our simulations allows us to compute the absolute error in a given season \(E_s\) from each estimation method and scenario as given by

@@

Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial found at \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}. After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T 2022). Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). The specifications for Settings A are more typical for stock assessment application (Thorson, James T 2022), and we therefore focus our discussion on these VAST results. Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.

Each biomass index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider. The columns in Table \ref{tab:scenarios} show the choices that define each scenario.

@@

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our 20-year biomass estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). The difference in variance is further illustrated by the median error values, with model-based estimates resulting in a median value of 0.235 while stratified mean values had a median error of 0.34. When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall absolute errors and standard deviations. Models with covariates had an average absolute error of 0.22 (median 0.19), compared to models that did not include covariate information having an average error of 0.46 (median 0.31), and stratified mean estimates produced an average error of 0.38 (median 0.34). These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide biomass estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined through yearly estimate:model ratio values, where we divide the yearly biomass estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly biomass values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).

Analysis of individual yearly estimate:model ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed similar patterns.

When the entire domain is sampled, the stratified mean produced an average yearly estimate:model ratio of 0.87 with a standard deviation of 0.12. Adding covariates in these scenarios brings the VAST estimate ratio closer to 1. Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in turn decreases the corresponding estimate ratio. For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10). With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19. As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.

@@

Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our biomass estimates for Yellowtail comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in \(67/96\approx 70\%\) of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved biomass estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. However, including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.

@@

To test the impact of shifting populations on biomass estimates, we simulated population data that included environmental forcing and compared biomass estimates derived from SRS to the known true values. We carried out simulations for 3 species on Georges Bank that took into account different population trends, temperature scenarios, noise in survey samples, and seasonal variation. We compared a design-based estimation approach (stratified mean) with a model-based approach that allows for the inclusion of environmental covariates (VAST) when the entire domain was sampled as well as when a reduced number of strata were included to simulate our species shifting into a new habitat. Our analysis included a comparison of absolute error values between 20-year estimates and consideration of the year-to-year variation displayed by each approach as well as the variation between each 20-year estimate.

Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true biomass trend. While VAST estimates tended to provide lower absolute error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 biomass estimates with larger error values than the largest stratified mean error. These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses. Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken in determining which diagnostic tools are used. For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model \textbf{cite Chris C paper}. The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data to account for factors such as environmental change. However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.

Seasonal differences in spatial biomass patterns in our population models led to seasonal differences in biomass estimates, with most estimates producing a higher error in the fall. This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 \& 14. Spawning in our population models concentrates much of each species in the spawning grounds towards the center of the domain where more samples occur. The populations are more spread out during the fall survey in weeks 37 \& 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled. We explored this through biomass estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.

@@

Of the \textasciitilde27\% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. This is in contrast to our Yellowtail results, where adding covariate information to VAST models always decreased the absolute error in the resulting biomass estimates. Although strata were excluded for both species, the decline in performance for just Cod and Haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists. We notice in Figure \ref{fig:strata-plot} that the strata excluded from sampling for Yellowtail still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate. On the other hand, the specific strata that were excluded from sampling for Haddock were located in the northern region of the domain, which means the covariate response for Haddock was not informed by any data from the coldest temperature range. This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also confounds the ability to project estimates into an unsampled region. Figure \ref{fig:CovResponse} depicts the impact on biomass estimates of including a covariate response for Haddock when both the average temperature and biomass were increasing over time. When projecting into the unsampled strata as seen Figure \ref{fig:CovResponse}, projecting biomass into the unsampled strata provides the wrong magnitude of biomass and also a different trend.

@@

The authors would like to thank Joe Langan for helping to develop the proposal that funded this work. Funding was provided by the Northeast Fisheries and Climate program at the Northeast Fisheries Science Center. Thank you to Jim Thorson for consulting with us about VAST settings and implementation. We are thanksful for Chris Cacciapaglia and Charles Adams for the time they spent with us sharing their experiences using VAST. We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-nefsc1218}{}}%
NEFSC. 2012. {54th Northeast Regional Stock Assessment Workshop (54th SAW) Assessment Report}. {Northeast Fish. Sci. Cent. Ref. Doc. 12-18; 604 p. Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.

\leavevmode\vadjust pre{\hypertarget{ref-nefsc1311}{}}%
NEFSC. 2013. {55th Northeast Regional Stock Assessment Workshop (55th SAW) Assessment Report}. {Northeast Fish. Sci. Cent. Ref. Doc. 13-11; 849 p. Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.

@@

@@

Table 1. Parameters used in all population models. SAW 1: (NEFSC 2012), SAW 2: (NEFSC 2013), SAW 3: not out yet???
